---
title: "Workshop Luxembourg Phenotypic Axes"
author: "Cynthia Sandor"
date: 'Last update: `r date()`'
output:
  html_document:
    fig_height: 5
    fig_width: 5
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
      toc_depth: 5
  pdf_document:
    toc: yes
---



```{r setup, include=FALSE}
library(phenix)
load('PPMI.Rdata')
```

## Using PHENIX

### Introduction

PHENIX (Dahl et al., 2016) use a Bayesian multiple-phenotype mixed model (MPMM), where the correlations between clinical phenotypes (Y) are decomposed into a genetic and a residual component with the following model: Y=U+e, where U represents the aggregate genetic contribution (whole genotype) to phenotypic variance and e is idiosyncratic noise.

As the estimation of maximum likelihood covariance estimates can become computationally expensive with increasing number of phenotypes, PHENIX uses a Bayesian low-rank matrix factorization model for the genetic term U such as: U = Sβ, in which β is can be used to estimate the genetic covariance matrix between phenotypes and S represents a matrix of latent components that each follow ~N (0,G) where G is the Estimate of Relatedness Matrix from genotypes.

The resulting latent traits (S) are used as phenotypic axes, each representing the severity of a number of non-independent clinical phenotypes.

To run PHENIX, we need: (1) Phenotypic matrix (P) (2) Kinship Matrix calculted from genotype from individual.
 
### Phenotypic matrix

The PPMI cohort (http://www.ppmi-info.org) was already described in detail (including PPMI protocol of recruitment and informed consent) by Marrek et al. (Parkinson Progression Marker, 2011).

We downloaded data from the PPMI database on September 2017 in compliance with the PPMI Data Use Agreement. We considered 472 newly-diagnosed typical Parkinson’s disease subjects: subjects with a diagnosis of Parkinson’s disease for two years or less and who are not taking Parkinson’s disease medications.

We used the baseline (t=0) of clinical assessments, described in detail in the Supplementary Table 2. We excluded any individual with > 5% of missing data (441 individuals included). 

We used the baseline following clinical assessement

```{r clinical assement ,echo=FALSE}
index_phen<-def_phen$Variable.Name %in% colnames(Y)
df<-def_phen[index_phen,]
DT::datatable(df, rownames = TRUE, caption = "Clinical Assessements", escape = FALSE, extensions = 'Buttons', options = list(dom = 'ftipr'))
```

### Kindship matrix

The kinship matrix G that is used by PHENIX to identify who’s related to each other, was created by using the GEMMA software (version 0.95) (Zhou and Stephens, 2012; Zhou et al., 2013; Zhou and Stephens, 2014) from genotype file under plink binary format. We pruned out SNPs in LD by using the following plink commands:

```
./plink - - bfile *plink binary files* - - indep-pairwise 50 5 0.2 - - out *output filename 1*
./plink - - bfile *plink binary files* - - exclude *output filename 1* - - make-bed - - out *output filename 2*
```

We then ran GEMMA with the following command:

```
./gemma -bfile *plink binary files of Parkinson’s disease cases with pruned SNPs* -gk 2 -o *output filename*
```


### Running PHENIX

We have two matrices of dimensions

```{r matrices dimension ,echo=FALSE}
print ("Dimension of the phenotypic matrix (nxp)")
dim(Y)
print ("Dimension of the Kinship matrix (nxn)")
dim(K)
```

Running PHENIX
The options quantnorm, scale and trim are important because they concern how PHENIX treats the data before generating the components:

Quantile normalizes phenotypes prior to imputation. While trim indicate to remove the outlier

```{r run phenix ,eval=TRUE,warning=FALSE,message=FALSE,error=FALSE}
out_phenix<-phenix (Y, K, quantnorm=TRUE, scale=TRUE, trim=FALSE)
```

The latent variables that correspond to phenotypic axes are saved in phenix_object$S. The dimension of this matrix is the same than phenotypic matrix.


```{r dimension of matrix of latent variables ,echo=FALSE}
print("Dimension of latent variables matrix (phenotypic axes)")
S<-out_phenix$S
print(dim(S))
```

## Phenotypic axis proprieties

### Fraction of the clinical variation

All axes don't explain the same proportion of the phenotypic variation. 

```{r proportion of phenotypic variation ,echo=FALSE}
sd_beta<-apply(out$beta[,c(1:33)], 1, sd)
var_expl<-cumsum((sd_beta)^2) / sum(sd_beta^2) 
barplot(var_expl[1:length(sd_beta)],names.arg=paste0("V",c(1:length(sd_beta))),las=2,cex.names = 0.8)
print ("3 first axes explain")
print(var_expl[3])
```

### First phenotypic axes are robust in term of clinical variables considered

We can remove randomly five clinical features
We observed a strong correlation (r > 0.8) between each of the two first phenotypic axes built with as few as 50% of the clinical variables and their respective original phenotypic axes, suggesting that these two axes are extremely robust in terms of the numbers of clinical variables considered

```{r remove randomly some clinical assessements ,echo=FALSE}
n_miss<-5
n_phen<-dim(Y)[2]
set.seed(123458)
list_random_phen<-sample(c(1:n_phen),5)
print("We remove the following clinical features")
print(colnames(Y)[list_random_phen])
Y_miss<-Y[,-list_random_phen]
print("New phenotypic matrix has as dimension")
dim(Y_miss)
```


```{r run phenix missing phenotypes,eval=TRUE,warning=FALSE,message=FALSE,error=FALSE}
print("Running PHENIX with this new matrix")
out_phenix_miss<-phenix (Y_miss, K, seed=8473, quantnorm=TRUE, scale=TRUE, trim=FALSE)
```

```{r examine correlation with two first axis,echo=FALSE}
plot(out_phenix$S[,1],out_phenix_miss$S[,1],xlab="Original Axis 1",ylab="Axis 1 build with missing phenotype")
plot(out_phenix$S[,2],out_phenix_miss$S[,2],xlab="Original Axis 2",ylab="Axis 2 build with missing phenotype")
```


### Each phenotypic axis represents a distinct set of clinical features

To interpret the clinical relevance of each phenotypic axis, we examined the correlation between individual clinical features and the phenotypic axes. 
To interpret these correlations it is important to know for each individuals clinical feature if a high score is associated with worsening symptoms or not.
For example a high score for UPDRS III correspond to worsening motors symptoms, while a high score for MoCA test reflect a better cognitive function for the patient. In consequence the interpretation of positive or negative correlation will depend on the clinical feature.

```{r correlation PPMI,echo=FALSE}

n_clinical<-length(colnames(Y))
n_axes<-2
list_axes<-c(1,2)
mm <- matrix(NA,n_clinical,n_axes)
for (i in 1:n_axes) {
  for (j in 1:n_clinical) {
    t<-cor.test(Y[,j],out_phenix[["S"]][,i])
    mm[j,i]<-as.numeric(round(t$estimate,3))
  }
}
df_mm<-as.data.frame(mm)
df_mm$Variable.Name<-colnames(Y)
df_mm_info<-merge(df_mm,def_phen,by=c("Variable.Name"))

df_mm_info<-df_mm_info[,c(1,4,2,3)]
colnames(df_mm_info)[1]<-"Clinical_Obs"
DT::datatable(df_mm_info, rownames = TRUE, caption = "Correlation Phenotypic Axes with Clinical Features", escape = FALSE, extensions = 'Buttons', options = list(dom = 'ftipr'))
```

```{r correlation plot PPMI,echo=FALSE}
library("reshape2")
library(ggplot2)

df_mm_info_s<-melt(df_mm_info,id=c("Clinical_Obs","Category"))
colnames(df_mm_info_s)[3]<-"Axis"
df_mm_info_s$Clinical_Obs<- factor(df_mm_info_s$Clinical_Obs, levels=unique(df_mm_info_s$Clinical_Obs[order(df_mm_info_s$Category)]))
ggplot(df_mm_info_s, aes(x=Clinical_Obs, y=value,fill=Category))+facet_grid(~Axis)+
 geom_bar(stat='identity',position=position_dodge()) + geom_bar(stat='identity',position=position_dodge()) +  theme(axis.text.x = element_text(angle = 90, hjust = 1))+coord_flip()

```



## Applications

Phenotypic axes are quantive measure of the clinical variation and can be use to explore the genetic underlying the clinical variation or identify new biomarkers such as Cerebrospinal fluid.
They measured for 120 individuals:

* Total tau
* Phosphorylated tau
* Amyloid beta peptide 1-42 (Aβ1-42)
* Alpha-synuclein


```{r four markers,echo=FALSE}
list_test<-unique(bio_marker_f$TESTNAME)
print (list_test)
```

We can examine the relation between these markers and phenotypic axes

```{r relation axes with csf markers,echo=FALSE}
df_axis<-data.frame(S[,c(1,2)])
df_axis$PATNO<-row.names(Y)
df_axis_bio_marker<-merge(df_axis,bio_marker_f[,c(1,6,7)],by=c("PATNO"))
colnames(df_axis_bio_marker)[2]<-"V1"
colnames(df_axis_bio_marker)[3]<-"V2"
df_axis_bio_marker_s<-melt(df_axis_bio_marker,id=c("PATNO","TESTNAME","TESTVALUE"))
colnames(df_axis_bio_marker_s)[4]<-"Axis"
ggplot(df_axis_bio_marker_s, aes(x=value, y=log( as.numeric(as.character(df_axis_bio_marker_s$TESTVALUE))),10) )+facet_grid(Axis~TESTNAME)+geom_point(aes(colour=factor(TESTNAME)))+ geom_smooth(method='lm',formula=y~x)+ylab("log10 concentration")
```

We find a significant relation between Aβ1-42 and phenotypic axes 
CSF Aβ1-42 was showed to predicts cognitive impairment in de novo PD patients. Phenotypic axis will enable to analysis patients from different cohorts with different clinical assessments.

```{r correlation between V1 and AB,echo=FALSE}
df_axis_bio_marker$TESTVALUE2<-as.numeric(as.character(df_axis_bio_marker$TESTVALUE))
x<-list_test[1]
cor.test(df_axis_bio_marker$V1[df_axis_bio_marker$TESTNAME==x],log(df_axis_bio_marker$TESTVALUE2[df_axis_bio_marker$TESTNAME==x],10))
```

